Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-08 csf-2017/paper.tex
Added FTLR
updated the theorem to a readable version
pass over intro and logrel

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control flow
integrity, local state encapsulation etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control flow integrity and
local state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) or rely on techniques like
machine code rewriting
(e.g.~\cite{wahbe_efficient_1993,abadi_control-flow_2005}), machine code
verification (e.g.~\cite{morrisett_system_1999}), virtual machines with a native
stack (e.g.~the JVM~\cite{lindholm_java_2014}) or
randomization~\cite{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors, intended to remediate these
limitations by presenting a better security model at the hardware level. They
are based on old ideas
(e.g.~\cite{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999}),
but have recently received renewd interest with particularly the CHERI project
proposing new ideas and tackling practical challenges like backwards
compatibility and realistic OS
support~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to enforce a
strict separation between numbers and capabilities, a kind of pointers carrying
some kind of authority. Memory capabilities carry the authority to read and/or
write to a range of memory locations. Additionally, the details vary but all
capability machines offer some form of \emph{object capabilities}, which
represent the authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g. the M-Machine's enter capabilities or CHERI's
sealed code/data pairs).

Unlike for commodity processors, the security primitives on capability machines
lend themselves well to the enforcement of properties like local state
encapsulation and control-flow integrity. Potentially, they will enable new
compilation schemes that enforce such properties in a way that is efficient but
also 100\% watertight (ideally evidenced by a mathematical proof, guaranteeing
that we do not end up in a new attack-defense arms race). However, there is
still a lot that needs to happen to attain this goal. For example, it is quite
non-trivial to devise a compilation scheme adapted to the details of a specific
language's notion of encapsulation (e.g. encapsulation of private member
variables in OO languages often behaves quite differently than encapsulation of
private state in ML-like languages). And even if such a scheme were defined, a
formal proof depends on a formalisation of the encapsulation provided by the
capability machine at hand.

A similar problem is the enforcement of control-flow integrity on capability
machines. An interesting approach is taken in CheriBSD. They drop the standard
contiguous C stack and split it into a central, trusted shadow stack, managed by
trusted call and return instructions\footnote{These call and return instructions
  are implemented by CheriBSD and special support from the CHERI processor makes
  them very fast to invoke.}, and disjoint, per-component
stacks~\citep{Watson2015Cheri}. For preventing illegal use of stack references,
the approach relies on \emph{local capabilities}, a type of capabilities offered
by CHERI that can be passed over a function invocation \emph{temporarily}, for
the duration of the invocation, and revoked afterwards. However, a lot of
details remain to be filled in (e.g. how to combine a trusted shadow stack with
(potentially) cross-domain function pointers in a scalable way?) and there is no
argument that this enforcement of CFI is watertight.

In this paper, we make progress towards the goal of efficient, provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item We formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard contiguous stack. It
  supports higher-order cross-component calls (e.g. (potentially)
  cross-component function pointers) and can be efficient assuming one
  additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem (capability-safety) that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed logical
  relation with recursively defined Kripke worlds. It is inspired by one used by
  \cite{Devriese:2016ObjCap} for capability-safety in a JavaScript-like lambda
  calculus, but has two novel technical characteristics. The natural CPS style
  of assembly code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal closure that
  has been previously used for lambda calculi. Perhaps more surprisingly, we
  show that the special nature of local capabilities can be elegantly expressed
  using a variant of an idea that has been used by
  \cite{Dreyer:jfp12} to express well-bracketedness
  guarantees in a lambda calculus: Kripke worlds with a weak and a strong future
  world relations. Interestingly, the two future world relations generally work
  very similarly to Dreyer et al.'s, except that they turn up in quite a
  different place in the logical relation.
\item We demonstrate the power of our results by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability-safety and our stack management discipline.
\end{itemize}

@@

The condition for capabilities with read permission is
$\readCond{}$. This condition makes sure the read capability can only
be used to read capability safe words. To make sure this is the case,
the region that governs the part of memory the capability grants
authority over must at least require all the words to be in the value
relation.
Generally speaking, global capabilities should only depend on
permanent regions. The function $\var{localityReg}(g,W)$ projects out
all active regions when $g$ is local, but when $g$ is global, then it
is just the permanent regions.

For capabilities with a write permission, $\writeCond{}$ must be
satisfied for the capabilities range of authority. Generally speaking,
write capabilities may be used to write any word, so the region that
governs the part of memory the write capability has authority over
must allow any capability safe word.
This is of course not entirely true because of the distinction between
write and permit-write-local capabilities. For write capabilities, the
region must require the contents of memory to be non-local and
capability safe words. For permit-write-local capabilities, the
non-local requirement is omitted.
The write condition uses $\var{localityReg}$ to pick an appropriate
region in the same way the read condition does.
The condition for the region to be address-stratified is there for
technical reasons.\lau{Dominique, you probably disagree with this, so
  please feel free to add something.}

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and
differs only slightly. Executable capabilities can be updated to point
anywhere in memory, so they must be capability safe as a program
counter no matter where they point in memory. Contrarily, the enter
capabilities are opaque and must be used from the address they point
to. They change permission when jumped to, so we require them to be
capability safe as a program counter when the permission is changed to
read-execute.
In both conditions, the capability must be capability safe as a
program counter certain future worlds. If the capability is global,
then it should be capability safe as a program counter in any private
future world. This means that it can only depend on permanent regions
because temporary regions may be revoked in a private future
world. On the other hand, local capabilities must only be capability
safe in public future worlds because they may depend on temporary
regions, so we cannot expect them to function correctly as program
counters if temporary regions are revoked.

@@

\subsection{Fundamental theorem of logical relations}

>>>>

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control flow
integrity, local state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control flow integrity and
local state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) or rely on techniques like
machine code rewriting
(e.g.,~\cite{wahbe_efficient_1993,abadi_control-flow_2005}), machine code
verification (e.g.,~\cite{morrisett_system_1999}), virtual machines with a native
stack (e.g.,~the JVM~\cite{lindholm_java_2014}) or
randomization~\cite{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

\emph{Capability machines} are a type of processors, intended to
remediate these limitations by presenting a better security model at
the hardware level. They are based on old ideas
(e.g.~\cite{Carter:1994:HSF:195473.195579,Dennis:1966:PSM:365230.365252,shapiro_eros:_1999}),
but have recently received renewed interest; in particular, the CHERI
project has proposed new ideas and ways of tackling practical
challenges like backwards compatibility and realistic OS
support~\cite{Watson2015Cheri,Woodruff:2014:CCM:2665671.2665740}. Capability
machines tag every word (in the register file and in memory) to
enforce a strict separation between numbers and capabilities, a kind
of pointers carrying some kind of authority. Memory capabilities carry
the authority to read and/or write to a range of memory
locations. Moreover, all capability machines
offer some form of \emph{object capabilities}, which represent the
authority to invoke a piece of code without exposing the code's
encapsulated private state (e.g., the M-Machine's enter capabilities or
CHERI's sealed code/data pairs).

In contrast to the situation for commodity processors, the security
primitives on capability machines lend themselves well to the
enforcement of properties like local state encapsulation and
control-flow integrity. Potentially, they will enable new compilation
schemes that enforce such properties in a way that is efficient but
also 100\% watertight (ideally evidenced by a mathematical proof,
guaranteeing that we do not end up in a new attack-defense arms
race). However, there is still a lot that needs to happen to attain
this goal. For example, it is quite non-trivial to devise a
compilation scheme adapted to the details of a specific language's
notion of encapsulation (e.g., encapsulation of private member
variables in OO languages often behaves quite differently than
encapsulation of private state in ML-like languages). And even if such
a scheme were defined, a formal proof depends on a formalisation of
the encapsulation provided by the capability machine at hand.

A similar problem is the enforcement of control-flow integrity on capability
machines. An interesting approach is taken in
CheriBSD~\citep{Watson2015Cheri}: It drops the standard
contiguous C stack and splits it into a central, trusted shadow stack, managed by
trusted call and return instructions\footnote{These call and return instructions
  are implemented by CheriBSD and special support from the CHERI processor makes
  them very fast to invoke.}, and disjoint, per-component
stacks. To prevent illegal use of stack references,
the approach relies on \emph{local capabilities}, a type of capabilities offered
by CHERI that can be passed over a function invocation \emph{temporarily}, for
the duration of the invocation, and revoked afterwards. However, a lot of
details remain to be investigated (e.g., how to combine a trusted shadow stack with
(potentially) cross-domain function pointers in a scalable way?) and there is no
argument that this enforcement of control-flow integrity is watertight.

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item We formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., (potentially)
  cross-component function pointers) and can be efficient assuming
  only \lars{can we say only here ?}
  one additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed
  logical relation with recursively defined Kripke worlds. It is
  inspired by one used in \cite{Devriese:2016ObjCap} for
  capability-safety in a JavaScript-like lambda calculus, but has two
  novel technical characteristics. The natural CPS style of assembly
  code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal
  closure that has been previously used for lambda calculi. Perhaps
  more surprisingly, we show that the special nature of local
  capabilities can be elegantly expressed using a variant of an idea
  that has been used by Dreyer et. al.~\cite{Dreyer:jfp12} to express
  well-bracketedness guarantees in a lambda calculus: Kripke worlds
  with both a weak and a strong future world relation. Interestingly,
  the two future world relations generally work very similarly to
  Dreyer et al.'s, except that they turn up in quite a different place
  in the logical relation.
\item We demonstrate the power of our results by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability-safety and our stack management discipline.
\end{itemize}

For reasons of space, som details and all proofs have been omitted;
please refer to the technical appendix~\cite{} for those.

@@

The condition for capabilities with read permission is
$\readCond{}$. It ensures that a read capability can only be used to
read capability safe words, by requiring that the region that governs
the part of memory that the capability grants authority over enforces
that all the words are in the value relation.  Generally speaking,
global capabilities should only depend on permanent regions. Therefore
we use the function $\var{localityReg}(g,W)$, which projects out all
active regions when $g$ is local, but only the permanent regions when
$g$ is global.

For a capability with a write permission, $\writeCond{}$ must be
satisfied for the capability's range of authority. Generally speaking,
write capabilities may be used to write any word, so the region that
governs the part of memory the write capability has authority over
must allow any capability safe word to be stored.
This is of course not entirely true because of the distinction between
write and permit-write-local capabilities. For write capabilities, the
region must require the contents of memory to be non-local (and
capability safe). For permit-write-local capabilities, the
non-local requirement is omitted.
The write condition uses $\var{localityReg}$ to pick an appropriate
region in the same way the read condition does.
The condition for the region to be address-stratified is there for
technical reasons.\lau{Dominique, you probably disagree with this, so
  please feel free to add something.}

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and
differ only slightly. Executable capabilities can be updated to point
anywhere in memory, so they must be capability safe as a program
counter no matter where they point in memory. In contrast, enter
\lars{``capability safe as a program counter'' is a bit hard to
  understand - why not say that this is a capability one can jump to ?}
capabilities are opaque and must be used from the address they point
to. They change permission when jumped to, so we require them to be
capability safe as a program counter when the permission is changed to
read-execute.
In both conditions, the capability must be capability safe as a
program counter in certain future worlds. If the capability is global,
then it should be capability safe as a program counter in any private
future world. This means that it can only depend on permanent regions
because temporary regions may be revoked in a private future
world. On the other hand, local capabilities must only be capability
safe in public future worlds because they may depend on temporary
regions, so we cannot expect them to function correctly as program
counters if temporary regions are revoked.

@@

\subsection{Fundamental theorem of logical relations}
\begin{theorem}
  \label{thm:ftlr}
  For all $n$, $\perm$, $\start$, $\addrend$, $\addr$, $\gl$, $W$  \\
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{\footnotesize{ and }} \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then
  \[
    \npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)
  \]
\end{theorem}
